### 2018-04-11 paper/abstract.tex
pass on first half of paper

\begin{abstract}
  % problem, context
  Recursion-free Constrained Horn Clauses (CHCs) are logic-programming
  problems that formulate verifying safety of programs with bounded
  iteration and recursion.
  %
  They both formulate bounded verification problems and are generated
  by solvers that attempt to verify safety of unbounded iterative or
  recursive systems.
  %
  Efficient solvers of recursion-free systems reduce the problem to
  solving a series of systems that can each be solved tractably.

  % contribution:
  In this paper, we define a novel tractable class of recursion-free
  systems, named \emph{Clause-Dependence Disjoint (CDD)} systems, that
  strictly generalizes tractable classes defined in previous work.
  %  
  We implemented a novel solver for recursion-free CHCs, named \sys,
  that reduces the problem of solving a given recursion-free system to
  solving an equivalent CDD system.
  %
  Our evaluation of \sys on standard benchmarks indicates that, in
  many practical cases, it performs significantly better than a
  state-of-the-art implementation.
\end{abstract}

>>>>

\begin{abstract}
  % problem, context
  Recursion-free Constrained Horn Clauses (CHCs) are logic-programming
  problems that can model safety properties of programs with bounded
  iteration and recursion.
  %
  In addition, many recursive CHC system solvers reduce inputs to a
  series of recursion-free CHC systems that can each be solved
  efficiently.

  % contribution:
  In this paper, we define a novel class of recursion-free systems,
  named \emph{Clause-Dependence Disjoint} (CDD), that
  generalizes classes defined in previous work.
  %
  The advantage of this class is that many systems are significantly
  smaller than equivalent systems in other classes which allows them
  to be solved more efficiently.
  %
  Based on the class of CDD systems, we implemented a CHC solver named
  \sys.  \sys~solves arbitrary CHC systems by reducing the input to a
  series of CDD systems.
  %
  Our evaluation indicates that \sys~outperforms state-of-the-art
  implementations in many practical cases.
\end{abstract}

### 2018-04-11 paper/introduction.tex
pass on first half of paper

% intro:
\section{Introduction}
\label{sec:intro}
% context, problem defn:
Many critical problems in program verification, such as the
verification of safety properties of recursive programs and
concurrent programs, can be reduced to solving systems of
\emph{Constrained Horn Clauses} (CHCs), a class of logic-programming
problems~\cite{bjorner13,flanagan03,rummer13a,rummer13b}.

% general introduce chc system
A CHC consists of a body of applications of uninterpreted relational
predicates, a constraint in a first-order theory, and a head
application of a relational-predicate.
%
A CHC system consists of a set of CHCs and a query relational
predicate.
% 
A solution of a system is an interpretation of each relational
predicate as a formula such that each clause body with predicate
symbols substituted with their interpretation entails the interpreted
clause head.
\QZ{This seems really out of place -- DAH}

% introduce the problem of solving recursion-free systems:
In this work we focus on the subclass of CHC systems which are known
as \emph{recursion-free}. In a recursion-free CHC system, the
derivation of each relational predicate will contain no references to
itself. \QZ{\DAH{I still think we need a small example here}} I.e., no
relational predicate is \emph{dependent} on itself.
%
Recursion-free CHC systems are an important subclass for two reasons.
%
First, recursion-free systems can be used to model safety properties
for hierarchical programs~\cite{lal-qadeer15,lal-qadeer-lahiri12},
i.e., programs with control branches and procedure calls, but only
bounded iteration and recursion.
%
Second (and most importantly), the task of solving a general CHC
system can reduce to the task of solving a sequence of recursion-free
systems.
%
Many \QZ{more references? Or remove many.} modern solvers generate
such a sequence where each subsystem is a bounded unwinding of the
input system. These solvers attempt to show that the solution to a
given subsystem generalizes to the original system or else generate a
larger subsystem for the next iteration~\cite{bjorner13}.
%
Such solvers are heavily reliant on the performance of the underlying
recursion-free solver since they invoke this underlying solver in an
iterative loop.
%
% current general techniques for solving recursion-free systems
Typically, even recursion-free CHC systems are not solved directly.
Instead, they are reduced to some even more specific subclass of CHC
system. For example, some approaches generate CHC systems whose
derivations are trees~\cite{bjorner13}.
\QZ{In general it seems we need many more references in the intro.}
%
A solver can dispatch such a system by issuing interpolation queries
to find a suitable definition for each relational predicate.
%
We refer to the subclass of CHC systems that can be solved by such
techniques as \emph{directly solvable}.

% Time complexity
In general, solving a recursion-free CHC system whose facts are
restricted to the theory of integer linear arithmetic is
co-NEXPTIME-complete~\cite{rummer13b}.
%
In contrast, solving a directly solvable system in the same theories
is co-NP~~\cite{rummer13b}.
%
Therefore, the performance of a CHC solver is highly reliant on the
order (i.e the number of predicates) of the directly solvable systems
it produces.

% current subclass
Previous work has introduced three classes of directly solvable
systems,
% derivation trees:
\textbf{(1)} tree systems~\cite{heizmann10,bjorner13,mcmillan14}, %
% disjunctive trees:
\textbf{(2)} body-disjoint systems, which represent multiple
derivation trees with a single disjunctive tree
system~\cite{rummer13a,rummer13b}, and
% linear systems:
\textbf{(3)} linear CHC systems, which can compactly represent
multiple linear derivations with a single system structured as a
Directed Acyclic Graph (DAG)~\cite{albarghouthi12a}.
%
The class of body-disjoint systems strictly contains the class of
derivation trees, and is independent of the class of linear systems
(i.e., the classes overlap, and neither contains the
other)~\cite{rummer13a,rummer13b}.

% contribution of this paper: CDD systems:
The first contribution of this paper is the introduction of a novel
class of directly solvable systems that is a superset of both
the body-disjoint subclass and the linear subclass.
%
In particular, the class consists of all recursion-free CHC systems
such that, for each clause $C$, the dependences of all distinct
relational predicates in the body of $C$ are disjoint \DAH{\QZ{Should
we illustrate what disjoin means here?}};
%
we thus refer to such systems as \emph{Clause-Dependent Disjoint
(CDD)} systems.
%
We demonstrate that the class of CDD systems is directly solvable
through a symbolic algorithm.
%

% a new solver
The second contribution of this paper is a solver for CHC systems,
named \sys.
%
Given a recursion-free system $S$, \sys reduces the problem
of solving $S$ to solving an equivalent CDD system
$S'$.
%
In the worst case, it is possible that the order of $S'$ may be
exponential in the order of $S$.
%
However, empirically we have found that the order of $S'$ is usually
close enough to the order of $S$ that \sys often outperforms the best
known CHC solvers.
%
Because the class of CDD systems is a superset of both body-disjoint
and linear systems, the order of an equivalent CDD system is always
less than or equal to that of systems in these subclasses.
%
The procedure implemented in \sys to construct $S'$ is a
generalization of existing procedures for generating compact
verification conditions of hierarchical
programs~\cite{flanagan01,lal-qadeer15}.
%
Given a general (possibly recursive) CHC system, \sys solves a sequence of
recursion-free systems.
%
Each subsystem is a bounded unwinding of the original system. \sys
attempts to combine the solutions of these recursion-free systems to
synthesize a soltuion of the original solution, as has been proposed
in previous work~\cite{rummer13b}.

% experience:
\QZ{need to change if need}
We implemented \sys within the \duality CHC solver~\cite{bjorner13},
which is implemented within the \zthree automatic theorem
prover~\cite{moura08}.
%
We evaluated the effectiveness of \sys on standard benchmarks drawn
from SVCOMP15~\cite{svcomp15}.
%
The results indicate that in a strong majority of practical cases,
\sys performs better than modern solvers.
%
Futhermore, the results indicate that combining the strengths of \sys
with that of other existing approaches (as discussed in
\autoref{sec:evaluation}) is a promising direction for the future of
CHC solving.

>>>>

% intro:
\section{Introduction}
\label{sec:intro}
% context, problem defn:
Many critical problems in program verification can be reduced to
solving systems of Constrained Horn Clauses (CHCs), a class of
logic-programming
problems~\cite{bjorner13,flanagan03,rummer13a,rummer13b}.
%
A CHC is a logical implication where the antecedent is a conjunction
of uninterpreted relational predicates with a fact in a background
theory and the consequent is an uninterpreted relational predicate. In
the context of a CHC we call the antecedent the body and the
consequent the head.  A CHC system is a set of CHCs.
%
The goal of a CHC solving problem is to find suitable interpretations
for each relational predicate such that each CHC is logically
consistent in isolation.

% introduce the problem of solving recursion-free systems:
In this work we focus on the subclass of CHC systems which are known
as \emph{recursion-free}. In a recursion-free CHC system, the
derivation of each relational predicate will contain no references to
itself. \QZ{\DAH{I still think we need a small example here}}
%
Recursion-free CHC systems are an important subclass for two reasons.
%
First, recursion-free systems can be used to model safety properties
for hierarchical programs~\cite{lal-qadeer15,lal-qadeer-lahiri12}
(programs with only bounded iteration and recursion).
%
Second and most importantly, the task of solving a general CHC
system can reduce to the task of solving a sequence of recursion-free
systems \QZ{reference}.

Many \QZ{more references? Or remove many.} modern solvers generate
a sequence such that each subsystem is a bounded unwinding of the
input system. These solvers either attempt to show that the solution to a
given subsystem generalizes to the original system or generate a
larger subsystem for the next iteration~\cite{bjorner13}.
%
Since such solvers inspect recursion-free systems in an iterative
loop, they are heavily reliant on the performance of the underlying
recursion-free solver.
%
% current general techniques for solving recursion-free systems
Typically, even recursion-free CHC systems are not solved directly.
Instead, they are reduced to a more specific subclass of
recursion-free CHC system.
%
These classes include those of
\emph{body disjoint} (or \emph{tree structure})
systems~\cite{heizmann10,bjorner13,mcmillan14,rummer13a,rummer13b} and
of \emph{linear} systems~\cite{albarghouthi12a}.
%
We will discuss these classes in \autoref{sec:overview} and
\autoref{sec:related-work}.
%
Such classes have proved practical because a solver can find
suitable definitions for relational predicates by issuing
\emph{interpolation queries}.
%
When a class of CHC systems can be solved by interpolation queries, we
say the class is \emph{directly solvable}.

% Time complexity
In general, solving a recursion-free CHC system whose facts are
restricted to the theory of linear integer arithmetic is
co-NEXPTIME-complete~\cite{rummer13b}.
%
In contrast, solving a directly solvable system with the same theories
is in co-NP~\cite{rummer13b}.
%
Therefore, the performance of a CHC solver is highly reliant on the
size of the directly solvable systems it produces.

% contribution of this paper: CDD systems:
The first contribution of this paper is the introduction of a novel
class of directly solvable systems that we refer to as
\emph{Clause-Dependence Disjoint} (CDD).
%
The class consists of all recursion-free CHC systems such that each
clause in the system has at most one dependency on the same relational
predicate.
%
The key characteristic of this class is that CDD systems are often
smaller than equivalent systems in other directly solvable classes,
and therefore solving recursion free systems by rewriting them in CDD
form is often less computationally expensive.

% a new solver
The second contribution of this paper is a solver for CHC systems,
named \sys.
%
Given a recursion-free system $S$, \sys reduces the problem
of solving $S$ to solving an equivalent CDD system
$S'$.
%
In the worst case, it is possible that the size of $S'$ may be
exponential in the size of $S$.
%
However, empirically we have found that the size of $S'$ is usually
close enough to the size of $S$ that \sys often outperforms the best
known CHC solvers.
%
The procedure implemented in \sys is a generalization of existing
techniques that synthesize compact verification conditions for
hierarchical programs~\cite{flanagan01,lal-qadeer15}.
%
Given a general (possibly recursive) CHC system, \sys solves a
sequence of recursion-free systems.
%
Each subsystem is a bounded unwinding of the original system. \sys
attempts to combine the solutions of these recursion-free systems to
synthesize a soltuion to the original problem, as has been proposed
in previous work~\cite{rummer13b}.

% experience:
We implemented \sys within the \duality CHC solver~\cite{bjorner13},
which is implemented within the \zthree automatic theorem
prover~\cite{moura08}.
%
We evaluated the effectiveness of \sys on standard benchmarks drawn
from SVCOMP15~\cite{svcomp15}.
%
The results indicate that in a strong majority of cases,
\sys performs better than modern solvers.
%
Futhermore, the results indicate that combining the strengths of \sys
with that of other existing approaches (as discussed in
\autoref{sec:evaluation}) is a promising direction for the future of
CHC solving.

### 2018-04-11 paper/overview.tex
pass on first half of paper

%first introduce the question
In \autoref{sec:running-ex}, we describe a recursion-free CHC system,
$\mcchc$, that formulates verifying the safety of a small program,
\cc{dblAbs} (\autoref{fig:multicall-code}).
%
In \autoref{sec:solve-ex}, we show that $\mcchc$ is a
\emph{Clause-Dependent Disjoint (CDD)} system and how \sys can solve it by
encoding it into binary interpolants.
%
In \autoref{sec:not-in}, we illustrate that $\mcchc$ is not in
directly solvable classes introduced in previous work.
%

@@

% include code of running example and some figure:
\begin{figure}[t]
  \centering
  \begin{floatrow}[2]
    \ffigbox[.26\textwidth] %
    { \caption{\cc{dblAbs}: an example hierarchical program.} %
      \label{fig:multicall-code} }
    { \input{code/dblAbs.java} }
    %
    \ffigbox[.7\textwidth] %
    {\caption{$\mcchc$ shown as a directed hypergraph.
        % 
        Each relational predicate is depicted as a graph node
        while each clause is represented by a hyperedge.
        Each hyperedge is labelled by the fact in the corresponding CHC\@.
        Each node has a valid corresponding interpretation,
        written in braces.
      } %
      \label{fig:ex-graph} }
    % { \includegraphics[width=\linewidth]{fig/ex-deps.pdf} }
      {\input{fig/example}}
  \end{floatrow}
\end{figure}
% walk through code of running example:
%
\cc{dblAbs} is a small procedure that doubles the absolute value of
its input and stores the result in \cc{res}.
% state the safety problem:
The program also asserts that \cc{res} is greater than or equal to $0$
before exiting.
% talk about solving Horn Clauses:
Verifying this assertion reduces to solving a recursion-free
CHC system over a set of uninterpreted predicates that represent
the control locations in \cc{dblAbs}.
%
In particular, one such system $\mcchc$ is
%
\begin{align}
\label{chc}
  % semantic constraint of dbl:
\cc{dbl}(\cc{x}, \cc{d}) &\gets \cc{d} = 2 * \cc{x} \\
  % semantic constraint of line 3:
  \cc{L}_4(\cc{n}, \cc{abs}) &\gets  \cc{abs} = 0 \\
  % semantic constraint of line 5 (then branch)
  \cc{L}_6(\cc{n}, \cc{abs}) &\gets \cc{L}_4(\cc{n}, \cc{abs}) \land \cc{n} \ge 0 \\
  % semantic constraint of line 6 (else branch)
  \cc{L}_8(\cc{n}, \cc{abs}) &\gets \cc{L}_4(\cc{n}, \cc{abs}) \land \cc{n} < 0 \\
  % assign branch
  \cc{L}_9(\cc{n}, \cc{abs'}) &\gets \cc{L}_6(\cc{n},\cc{abs}) \land \cc{abs'} = \cc{n} \\
   \cc{L}_9(\cc{n}, \cc{abs'}) &\gets \cc{L}_8(\cc{n},\cc{abs}) \land \cc{abs'} = -\cc{n} \\
  % semantic constraint of main procedure:
   \cc{main}(\cc{n},\cc{res}) &\gets \cc{L}_9(\cc{n}, \cc{abs'}) \land
                                     \cc{dbl}(\cc{x}, \cc{d})
    \land \cc{abs'} = \cc{x} \land \cc{res} = \cc{d} \\
  %
    \bot &\gets \cc{main}(\cc{n}, \cc{res}) \land \cc{res} < 0 
\end{align}
%
While $\mcchc$ has been presented as the result
of a translation from hierarchical program \cc{dblAbs}, \sys is
purely a solver for CHC systems: it does not require access to the
concrete representation of a program, or for a given CHC system to be
the result of translation from a program at all.

% solving the running example
\subsection{$\mcchc$ as a Clause-Dependent Disjoint System}
\label{sec:solve-ex}
% state it is CDD
The recursion-free CHC system $\mcchc$ is a
\emph{Clause-Dependent-Disjoint} (CDD) system.
% state the defn of CDD:
A CDD system is a recursion-free CHC system such that
the dependencies between predicates in the system obey certain rules.
%
One predicate is directly dependent on another if the second appears
in the body of a clause where the first is the head.
%
Based on this, we say that a clause is dependence disjoint if
\textbf{(1)} no two predicates in the body share any transitive
dependencies on other predicates and \textbf{(2)} no two predicates in
the body are the same.
% give a example
As an example, clause $(7)$ is dependence disjoint. Clause $(7)$ uses
both $\cc{L}_9$ and $\cc{dbl}$ in its body.
%
The transitive dependency of $\cc{L}_9$ is the set
$\{\cc{L}_4,\cc{L}_6,\cc{L}_8\}$, while the transitive dependency of
\cc{dbl} is the empty set.
%
Therefore, their transitive dependencies are disjoint:
$\{\cc{L}_4,\cc{L}_6,\cc{L}_8\} \cap \varnothing = \varnothing$.
%
All other clauses in $\mcchc$ have at most one uninterpreted predicate
in the body, so they are trivially disjoint dependent.
%
When each clause in a CHC system is dependence disjoint, it is a CDD
system. Therefore $\mcchc$ is a CDD system.
%
The fact that $\mcchc$ is CDD corresponds to the fact that there no
statements executed more than once within one execution path.
\DAH{elaborate}
%
The formal definition of CDD and its key properties are given in
\autoref{sec:CDD-defn}.
The formal definition of transitive dependency is given in\QZ{reference}.
%

% restate defn of CDD systems:
\sys solves CDD systems directly by issuing a binary interpolation
query for each uninterpreted predicate in topological order.
%
Each uninterpreted predicate $P$ has a valid interpretation which can
be computed by interpolating
\textbf{(1)} the \emph{pre}-formula, constructed from the solutions of
each uninterpreted predicate $P$ directly depends on and constraints
in clauses where $P$ is the head and
\textbf{(2)} the \emph{post}-formula, constructed from
all clauses where the head transitively depends on $P$.
%

We will walk through the construction of an interpolation query for
$\cc{L}_9$ by hand.  By the time \sys reaches $\cc{L}_9$ it will
already have solutions for $\cc{L}_4$, $\cc{L}_6$, $\cc{L}_8$.
Possible interpretations of these predicates are shown in
\autoref{fig:ex-graph}.
%
A solution for \cc{L9} can be constructed by interpolating \textbf{(1)}
the \emph{pre}-formula, constructed as solutions generated for \cc{L4}, \cc{L6}
the constraints in clauses (5) and (6) and boolean variables%
\textbf{(2)} the \emph{post}-formula constructed from constraints in
$\mcchc$ that all constraints after \cc{L9}, constraints for \cc{dbl}
and boolean variables.
%
In particular, the \emph{pre}-formula for \cc{L9} is 
\begin{gather}
  ((b_{L_8}\land \cc{abs'}= -\cc{n})
    \lor (b_{L_6} \land \cc{abs}=\cc{n}))
  \land (\neg b_{L_6} \lor \cc{n} \ge 0)
    \land (\neg b_{L_8} \lor \cc{n}<0)
\end{gather}
%
, $b_{L8}$ and $b_{L6}$ are boolean variables corresponding to predicates \cc{L8} and \cc{L6}.
%
The \emph{post}-formula for \cc{L9} is
\begin{gather}
  (b_{main} \land \cc{res'} < 0 )
    \land (\neg b_{main} \lor (b_{dbl} \land \cc{abs'}= \cc{x} \land \cc{res}=\cc{d} ))
  \land (\neg b_{dbl} \lor \cc{d} = 2*\cc{x})
\end{gather}
, $b_{main}$ and $b_{dbl}$ are boolean variables corresponding to
predicates \cc{main} and \cc{dbl}.
%
A detailed and formal description for the procedure of solving a CDD system is given in
\autoref{sec:solve-cdd}.

\subsection{$\mcchc$ is not in other system}
\label{sec:not-in}
%
There are three classes of directly solvable systems has been introduced
in previous work(as described in \autoref{sec:intro}).
%
However, $\mcchc$ does not belong to any of these three classes.
%

% tree solver 
In particular, previous work has introduced tree structure systems\cite{bjorner13,heizmann10},
which each uninterpreted predicate in the system is the head of at most one clause.
%
Such systems do not include $\mcchc$, because predicate \cc{L9} is the head 
of clause (5) and (6).
%
CHC solvers based on tree structure system has to reduce solving $\mcchc$ to
solving two tree strucutre systems that contains by $\mcchc$, that one corresponds
to the tree branch and the other corresponds to the false branch.
% not that bad 
There are works has been proposed which in some cases enumerating 
and solving all derivation trees 
of a given recursion-free
system~\cite{mcmillan14}.
%
However, such solvers may enumerate all derivations tree of a given system
in the worst case.

% Body-Disjoint:
Previous work has introduced \emph{body-disjoint}
systems~\cite{rummer13a,rummer13b}, 
which each uninterpreted predicate shows in the body of at most one
clause, and shows in the body of such a clause at most once;
%
However, such systems cannot model a program with multiple
control paths that share a common subpath, typically modeled as a CHC
system with a uninterpreted predicate that occurs in the body of multiple clauses.
%
$\mcchc$ is not a body-disjoint system because the uninterpreted
predicates \cc{L4} shows in the body of both clauses (3) and (4).
%
Such a solver, given $\mcchc$, generates a body-disjoint system
consisting of two copies of the sub-systems with head relational
predicates \cc{L4}.

%linear systems:
Previous work also introduced linear system~\cite{albarghouthi12a}, 
which each clause's body has at most one uninterpreted predicate.
%
However, such systems cannot directly model the control flow of a
program that contains procedure calls.
%
$\mcchc$ is not a linear system because the body of clause (7) has two
predicates \cc{L9} and \cc{dbl}.
%
CHC solvers that solve linear systems, $\mcchc$, effectively inline constraints on
relational predicates that occur in non-linear clauses~\cite{albarghouthi12b}.
%
Such approaches can, in general, generate an inlined system that is
exponentially larger than a given system.

%% talk about expansion:
The class of CDD systems strictly contains these three classes, 
but does not contain all
recursion-free systems.
%
E.g., a recursion-free system that models \cc{dblAbs} 
call \cc{dbl} twice on \cc{abs} in sequence would not
be CDD.
%
\sys, given an arbitrary recursion-free CHC system $S$,
solves it by generating an equivalent CDD system $S'$ and
solving $S'$ directly.
%
In general, $S'$ may have size exponential in the size of
$S$.
%
In practice, a suitable equivalent system can be generated using
heuristics analogous to those used to generate compact verification
conditions of hierarchical programs~\cite{flanagan01,lal-qadeer15}.
%
A general procedure for constructing a CDD expansion of a given CHC
system is given in \autoref{app:cons-cdd}.

>>>>

%first introduce the question
In \autoref{sec:running-ex}, we describe a recursion-free CHC system,
$\mcchc$, that models the safety of a small program,
\cc{dblAbs} (\autoref{fig:multicall-code}).
%
In \autoref{sec:solve-ex}, we show that $\mcchc$ is a CDD system and
how \sys can solve it by encoding it into binary interpolants.
%
In \autoref{sec:not-in}, we illustrate that $\mcchc$ is not in
directly solvable classes introduced in previous work.
%

@@

% include code of running example and some figure:
\begin{figure}[t]
  \centering
  \begin{floatrow}[2]
    \ffigbox[.26\textwidth] %
    { \caption{\cc{dblAbs}: an example hierarchical program.} %
      \label{fig:multicall-code} }
    { \input{code/dblAbs.java} }
    %
    \ffigbox[.7\textwidth] %
    {\caption{$\mcchc$ as a directed hypergraph.
        % 
        Each relational predicate is depicted as a graph node
        while each clause is represented by a hyperedge.
        Each hyperedge is labelled by the fact in the corresponding CHC\@.
        Each node has a valid corresponding interpretation,
        written in braces.
      } %
      \label{fig:ex-graph} }
    % { \includegraphics[width=\linewidth]{fig/ex-deps.pdf} }
      {\input{fig/example}}
  \end{floatrow}
\end{figure}
% walk through code of running example:
%
\cc{dblAbs} is a small procedure that doubles the absolute value of
its input and stores the result in \cc{res}.
% state the safety problem:
The program also asserts that \cc{res} is greater than or equal to $0$
before exiting.
% talk about solving Horn Clauses:
Verifying this assertion reduces to solving a recursion-free
CHC system over a set of uninterpreted predicates that represent
the control locations in \cc{dblAbs}.
%
In particular, one such system $\mcchc$ is
%
\begin{align}
\label{chc}
  % semantic constraint of dbl:
\cc{dbl}(\cc{x}, \cc{d}) &\gets \cc{d} = 2 * \cc{x} \\
  % semantic constraint of line 3:
  \cc{L}_4(\cc{n}, \cc{abs}) &\gets  \cc{abs} = 0 \\
  % semantic constraint of line 5 (then branch)
  \cc{L}_6(\cc{n}, \cc{abs}) &\gets \cc{L}_4(\cc{n}, \cc{abs}) \land \cc{n} \ge 0 \\
  % semantic constraint of line 6 (else branch)
  \cc{L}_8(\cc{n}, \cc{abs}) &\gets \cc{L}_4(\cc{n}, \cc{abs}) \land \cc{n} < 0 \\
  % assign branch
  \cc{L}_9(\cc{n}, \cc{abs'}) &\gets \cc{L}_6(\cc{n},\cc{abs}) \land \cc{abs'} = \cc{n} \\
   \cc{L}_9(\cc{n}, \cc{abs'}) &\gets \cc{L}_8(\cc{n},\cc{abs}) \land \cc{abs'} = -\cc{n} \\
  % semantic constraint of main procedure:
   \cc{main}(\cc{n},\cc{res}) &\gets \cc{L}_9(\cc{n}, \cc{abs'}) \land
                                     \cc{dbl}(\cc{x}, \cc{d})
    \land \cc{abs'} = \cc{x} \land \cc{res} = \cc{d} \\
  %
    \bot &\gets \cc{main}(\cc{n}, \cc{res}) \land \cc{res} < 0 
\end{align}
%
While $\mcchc$ has been presented as the result
of a translation from \cc{dblAbs}, \sys is
purely a solver for CHC systems: it does not require access to the
concrete representation of a program, or for a given CHC system to be
the result of translation from a program at all.

% solving the running example
\subsection{$\mcchc$ as a Clause-Dependence Disjoint System}
\label{sec:solve-ex}
% state it is CDD
The recursion-free CHC system $\mcchc$ is a
\emph{Clause-Dependence-Disjoint} (CDD) system.
% state the defn of CDD:
A CHC system can be classified as CDD when the clauses obey
rules based on dependencies between the predicates.
%
One predicate is directly dependent on another if the second appears
in the body of a clause where the first is the head.
%
Based on this, we say that a clause is dependence disjoint if
\textbf{(1)} no two predicates in the body share any transitive
dependencies on other predicates and \textbf{(2)} no two predicates in
the body are the same.
% give a example
As an example, clause $(7)$ is dependence disjoint. Clause $(7)$ uses
both $\cc{L}_9$ and $\cc{dbl}$ in its body.
%
The transitive dependency of $\cc{L}_9$ is the set
$\{\cc{L}_4,\cc{L}_6,\cc{L}_8\}$ while the transitive dependency of
\cc{dbl} is the empty set.
%
Therefore, their transitive dependencies are disjoint:
$\{\cc{L}_4,\cc{L}_6,\cc{L}_8\} \cap \varnothing = \varnothing$.
%
All other clauses in $\mcchc$ have at most one uninterpreted predicate
in the body, so they are trivially disjoint dependent.
%
When each clause in a CHC system is dependence disjoint, it is a CDD
system. Therefore $\mcchc$ is a CDD system.
%
%
The formal definition of CDD and its key properties are given in
\autoref{sec:CDD-defn}.
The formal definition of transitive dependency is given in\QZ{reference}.
%

% restate defn of CDD systems:
\sys solves CDD systems directly by issuing a binary interpolation
query for each uninterpreted predicate in topological order.
%
Each uninterpreted predicate $P$ has a valid interpretation which can
be computed by interpolating
\textbf{(1)} the \emph{pre}-formula, constructed from clauses where
$P$ is the head and
\textbf{(2)} the \emph{post}-formula, constructed from
all clauses where the head transitively depends on $P$.
%

We will walk through the construction of an interpolation query for
$\cc{L}_9$ by hand.  By the time \sys reaches $\cc{L}_9$ it will
already have solutions for $\cc{L}_4$, $\cc{L}_6$, $\cc{L}_8$.
Possible interpretations of these predicates are shown in
\autoref{fig:ex-graph}.
%
A solution for $\cc{L}_9$ can be constructed by interpolating
the appropriate pre and post formulas.
%
The pre-formula is constructed from the bodies of clauses where
$\cc{L}_9$ is the head. Each relational
predicate, $P$,
is replaced by a corresponding boolean indicator variable, $\cc{b}_P$, and each
boolean indicator variable implies the solution for its predicate.
%
In particular, the pre-formula for $\cc{L}_9$ is constructed from
clauses (5) and (6):
\begin{gather}
  ((\cc{b}_{\cc{L}_6} \land \cc{abs'} = \cc{n})
  \lor
  (\cc{b}_{\cc{L}_8} \land \cc{abs'} = -\cc{n}))
  \land
  (\cc{b}_{\cc{L}_6} \Rightarrow \cc{n} \ge 0)
  \land
  (\cc{b}_{\cc{L}_8} \Rightarrow \cc{n} < 0)
\end{gather}
%
The post-formula is constructed from clauses that transitively depend
on $\cc{L}_9$. Again, we replace relational predicates by
corresponding boolean indicators. However, we omit the boolean indicator for
$\cc{L}_9$. The post-formula is composed from clauses (1), (7), and
(8):
\begin{gather}
  (\cc{b}_{\cc{main}} \land \cc{res'} < 0)
  \land
  (\cc{b}_{\cc{main}} \Rightarrow
    (\cc{b}_{\cc{dbl}}
    \land \cc{abs'}= \cc{x}
    \land \cc{res}=\cc{d} ))
  \land (\cc{b}_{\cc{dbl}} \Rightarrow \cc{d} = 2*\cc{x})
\end{gather}
%
Interpolating the pre and post formulas yields a valid intepretation
of $\cc{L}_9$: $\cc{abs'} \geq 0$.
The procedure for solving a CDD system is described in formal detail
in \autoref{sec:solve-cdd}.

\subsection{$\mcchc$ is not in other recursion-free classes}
\label{sec:not-in}
%
In this section, we show that $\mcchc$ cannot be placed in other
known classes of recursion-free CHC systems. Specifically, we will
discuss body-disjoint systems and linear systems.
%

% tree solver
Tree structure
systems~\cite{mcmillan14,bjorner13,heizmann10,rummer13a,rummer13b} are
class of recursion-free CHC system where each uninterpreted predicate
appears in the body of at most one clause (and appears in such a
clause exactly once).
%
Such systems cannot model a program with multiple control paths that
share a common subpath, typically modeled as a CHC system with a
uninterpreted predicate that occurs in the body of multiple clauses.
%
$\mcchc$ is not a body-disjoint system because the uninterpreted
predicates $\cc{L}_4$ shows in the body of both clause (3) and clause
(4).
%
In order to find an interpretation for $\mcchc$, a solver which uses
tree-structure systems would have to create a fresh copy of
$\cc{L}_4$, as well copies for all relational predicates that
$\cc{L}_4$ depends on.

%linear systems:
Previous work also introduced linear system~\cite{albarghouthi12a},
where each clause's body has at most one uninterpreted predicate.
%
However, such systems cannot directly model the control flow of a
program that contains procedure calls.
%
$\mcchc$ is not a linear system because the body of clause (7) has two
predicates \cc{L9} and \cc{dbl}.
%
CHC solvers that solve linear systems effectively inline the
constraints for relational predicates that occur in non-linear
clauses~\cite{albarghouthi12b}.
%
In the case of $\mcchc$, inlining the constraints of $\cc{dbl}$ is
efficient, but in general such approaches can generate systems that
are exponentially larger than the input.

%% talk about expansion:
\begin{figure}
  \input{fig/hierarchy}
\end{figure}

The class CDD of CDD systems is a superclass of the classes of
body-disjoint and linear systems.
%
% The fact that $\mcchc$ is CDD corresponds to the fact that there no
% statements executed more than once within one execution path.
% \DAH{elaborate}
%
E.g., a recursion-free system that models \cc{dblAbs} 
call \cc{dbl} twice on \cc{abs} in sequence would not
be CDD.
%
\sys, given an arbitrary recursion-free CHC system $S$,
solves it by generating an equivalent CDD system $S'$ and
solving $S'$ directly.
%
In general, $S'$ may have size exponential in the size of
$S$.
%
In practice, a suitable equivalent system can be generated using
heuristics analogous to those used to generate compact verification
conditions of hierarchical programs~\cite{flanagan01,lal-qadeer15}.
%
A general procedure for constructing a CDD expansion of a given CHC
system is given in \autoref{app:cons-cdd}.
